---
title: "HOF MODEL"
format: html
editor: visual
---

## Libraries

```{r}
library(dplyr)
library(stringr)
library(janitor)
library(Lahman)
library(tidymodels)
library(readxl)


```

## Simplify AQCS Names

```{r}
AQCS <- read_excel("All Qualified Catchers Stats.xlsx") %>%
  clean_names() %>%
  mutate(
    name_clean = name %>%
      str_replace_all("\\.", "") %>%   
      str_replace_all(",", "") %>%     
      str_replace_all("'", "") %>%
      str_replace_all("-", " ") %>%    
      str_squish() %>%      
      str_to_lower()
  )

```

## Lahman Name Cleaning

```{r}
people_lookup <- Lahman::People %>%
  as_tibble() %>%
  transmute(
    playerID,
    name_clean = paste(nameFirst, nameLast) %>%
      str_replace_all("\\.", "") %>%
      str_replace_all(",", "") %>%
      str_replace_all("'", "") %>%
      str_replace_all("-", " ") %>%
      str_squish() %>%
      str_to_lower()
  )
```

## Join AQCS and Lahman

```{r}
AQCS <- AQCS %>%
  left_join(people_lookup, by = "name_clean")

manual_name_map <- tibble::tribble(
  ~name,                  ~playerID,
  "A.J. Pierzynski",      "pierzaj01",
  "B.J. Surhoff",         "surhobj01",
  "Billy Sullivan Jr.",   "sullibi04",
  "Billy Sullivan Sr.",   "sullibi03",
  "Chief Meyers",         "meyerch01",
  "Chief Zimmer",         "zimmech01",
  "Christian Vázquez",    "vazquch01",
  "Eddie Taubensee",      "taubeed01",
  "Gary Sánchez",         "sanchga02",
  "J.T. Realmuto",       "realmjt01",
  "Martín Maldonado",     "maldoma01",
  "Sandy Alomar Jr.",     "alomasa02",
  "Tony Peña Sr.",        "penato01"
)

AQCS <- AQCS %>%
  left_join(manual_name_map, by = "name", suffix = c("", "_manual")) %>%
  mutate(playerID = coalesce(playerID, playerID_manual)) %>%
  select(-playerID_manual) %>%
  filter(name != "Frank Duncan")


'
###CODE USED IN PROCESS###
#How many matched?
AQCS %>%
  summarise(
    total = n(),
    matched = sum(!is.na(playerID)),
    match_rate = matched / total
  )

unmatched_names <- AQCS %>%
  filter(is.na(playerID)) %>%
  distinct(name) %>%
  arrange(name)

unmatched_names
"
A.J. Pierzynski		
B.J. Surhoff		
Billy Sullivan Jr.		
Billy Sullivan Sr.		
Chief Meyers		
Chief Zimmer		
Christian Vázquez		
Eddie Taubensee		
Frank Duncan		
Gary Sánchez
"
find_people <- function(pattern) {
  People %>%
    as_tibble() %>%
    transmute(
      playerID,
      full_name = paste(nameFirst, nameLast),
      nameFirst,
      nameLast
    ) %>%
    filter(str_detect(str_to_lower(full_name), str_to_lower(pattern))) %>%
    arrange(nameLast, nameFirst)
}

find_people("Peña")

Lahman::People |> dplyr::filter(playerID == "penato01")
'

```

## Add All-Star Apps to AQCS

```{r}
allstar_counts <- Lahman::AllstarFull %>%
  as_tibble() %>%
  count(playerID, name = "all_star_selections")

AQCS <- AQCS %>%
  left_join(allstar_counts, by = "playerID") %>%
  mutate(all_star_selections = coalesce(all_star_selections, 0L))

###Sanity Checks Below###
'
AQCS %>% select(name, playerID, all_star_selections) %>% arrange(desc(all_star_selections)) %>% head(15)
AQCS %>% filter(str_detect(name, "Molina")) %>% select(name, all_star_selections)
'

```

## Add Gold Gloves, Silver Sluggers, and MVPs

```{r}
awards <- Lahman::AwardsPlayers %>%
  as_tibble()

# Gold Gloves (career count)
gg_counts <- awards %>%
  filter(awardID == "Gold Glove") %>%
  count(playerID, name = "gg")

# Silver Sluggers (career count)
ss_counts <- awards %>%
  filter(awardID == "Silver Slugger") %>%
  count(playerID, name = "ss")

# MVP awards won (career count)
mvp_awards <- awards %>%
  filter(awardID == "Most Valuable Player") %>%
  count(playerID, name = "mvp_awards")

AQCS <- AQCS %>%
  left_join(gg_counts,  by = "playerID") %>%
  left_join(ss_counts,  by = "playerID") %>%
  left_join(mvp_awards, by = "playerID") %>%
  mutate(
    gg = coalesce(gg, 0L),
    ss = coalesce(ss, 0L),
    mvp_awards = coalesce(mvp_awards, 0L)
  )

'
### CODE USED IN PROCESS ###
names(Lahman::AwardsSharePlayers)
head(Lahman::AwardsSharePlayers)

Lahman::AwardsSharePlayers %>%
  distinct(awardID) %>%
  arrange(awardID)

### Sanity Checks Below
AQCS %>%
  select(name, all_star_selections, gg, ss, mvp_awards) %>%
  arrange(desc(mvp_awards), desc(all_star_selections)) %>%
  head(20)
'


```

## Create HOF label

```{r}

cutoff_last_mlb_year <- 2012

hof_lookup <- Lahman::HallOfFame %>%
  as_tibble() %>%
  group_by(playerID) %>%
  summarise(
    inducted_any = any(inducted == "Y", na.rm = TRUE),
    .groups = "drop"
  )

last_mlb_year_lookup <- Lahman::Batting %>%
  as_tibble() %>%
  group_by(playerID) %>%
  summarise(last_mlb_year = max(yearID, na.rm = TRUE), .groups = "drop")

AQCS <- AQCS %>%
  left_join(hof_lookup, by = "playerID") %>%
  left_join(last_mlb_year_lookup, by = "playerID") %>%
  mutate(
    inducted_any = coalesce(inducted_any, FALSE),
    hof_resolved = inducted_any | (!is.na(last_mlb_year) & last_mlb_year <= cutoff_last_mlb_year),
    hof = case_when(
      inducted_any ~ 1L,
      hof_resolved ~ 0L,
      TRUE ~ NA_integer_
    )
  )
'
### Sanity Checks Below
# how many got labeled
AQCS %>%
  summarise(
    total = n(),
    labeled = sum(!is.na(hof)),
    labeled_rate = labeled / total,
    hof_yes = sum(hof == 1, na.rm = TRUE),
    hof_no  = sum(hof == 0, na.rm = TRUE)
  )
'
```

## Remove Unlabeled Players (recent retirees)

```{r}
AQCS_model <- AQCS %>%
  filter(!is.na(hof))

AQCS_holdout <- AQCS %>%
  filter(is.na(hof))

'
###Sanity Check Below
nrow(AQCS_model) + nrow(AQCS_holdout) == nrow(AQCS)

View(AQCS_model)
'
```

## Pick Model Variables

```{r}
model_features <- c(
  "war",
  "pa",
  "g",
  "w_rc",
  "w_oba",
  "def",
  "fielding",
  "positional",
  "all_star_selections",
  "gg",
  "ss",
  "mvp_awards"
)

AQCS_model_ready <- AQCS %>%
  filter(!is.na(hof)) %>%        # resolved only
  select(all_of(model_features), hof, last_mlb_year)

'
##Sanity Check Below
glimpse(AQCS_model_ready)
'



```

## Split Data

```{r}
train_cutoff_year <- 2005

train_data <- AQCS_model_ready %>%
  filter(last_mlb_year <= train_cutoff_year) %>%
  select(-last_mlb_year)

test_data <- AQCS_model_ready %>%
  filter(last_mlb_year > train_cutoff_year) %>%
  select(-last_mlb_year)

'
###IMPORTANT Sanity Checks Below
# Row counts
nrow(train_data)
nrow(test_data)

# Class balance
table(train_data$hof)
table(test_data$hof)

# Missing values check
sapply(train_data, function(x) sum(is.na(x)))
'
```

## Build the recipe for training (no training yet)

```{r}

#seed 4 for yadi himself
set.seed(4)

# Ensure outcome is a factor for classification
train_data <- train_data %>%
  mutate(hof = factor(hof, levels = c(0, 1)))

test_data <- test_data %>%
  mutate(hof = factor(hof, levels = c(0, 1)))

# Recipe: impute + scale
hof_recipe <- recipe(hof ~ ., data = train_data) %>%
  step_impute_median(all_predictors()) %>%
  step_normalize(all_predictors())

# Model: regularized logistic regression
hof_model <- logistic_reg(
  penalty = tune(),
  mixture = 0 
) %>%
  set_engine("glmnet")

# Workflow
hof_workflow <- workflow() %>%
  add_recipe(hof_recipe) %>%
  add_model(hof_model)

'
###Sanity Checks
table(train_data$hof)
table(test_data$hof)
sum(is.na(train_data))
sum(is.na(test_data))
'

```

## Overfitting Adjustments

```{r}

set.seed(4)

hof_folds <- vfold_cv(
  train_data,
  v = 5,
  repeats = 10,
  strata = hof
)


# Penalty grid (log scale works best)
penalty_grid <- grid_regular(
  penalty(range = c(-4, 1)),  # 10^-4 to 10^1
  levels = 20
)

# Tune
hof_tune <- tune_grid(
  hof_workflow,
  resamples = hof_folds,
  grid = penalty_grid,
  metrics = metric_set(roc_auc)
)

# Select best penalty
best_penalty <- select_best(hof_tune, metric = "roc_auc")

best_penalty

```

## Train Data, Fit Model

```{r}
set.seed(4)

final_workflow <- finalize_workflow(hof_workflow, best_penalty)

final_fit <- fit(final_workflow, data = train_data)

```

## Evaluate (once!) on the test set

```{r}
test_pred <- predict(final_fit, new_data = test_data, type = "prob") %>%
  bind_cols(test_data %>% select(hof)) %>%
  mutate(
    .pred_class = factor(
      if_else(.pred_1 >= 0.5, "1", "0"),
      levels = c("0", "1")
    )
  )

metrics <- metric_set(roc_auc, accuracy, sens, spec)

metrics(test_pred, truth = hof, estimate = .pred_class, .pred_1)

conf_mat(
  test_pred,
  truth = hof,
  estimate = .pred_class
)

### Model looks good to move on

```

## Create new Dataset with all labeled players

```{r}
predict_data <- AQCS %>%
  filter(is.na(hof))

```

## Reiterate Workflow

```{r}
final_workflow <- finalize_workflow(hof_workflow, best_penalty)

```

## Train model (once!)

```{r}
final_fit <- fit(final_workflow, data = train_data)
```

## Predict for unlabeled players

```{r}
predict_results <- predict(final_fit, new_data = predict_data, type = "prob") %>%
  bind_cols(predict_data %>% select(name, playerID)) %>%
  rename(p_hof = .pred_1) %>%
  arrange(desc(p_hof))

```

## RESULTS

```{r}
###YADI SPECIFICALLY
predict_results %>%
  filter(str_detect(name, "Molina")) %>%
  select(name, playerID, p_hof)

###Likelihood of catchers who retired after 2012
predict_results %>%
  select(name, playerID, p_hof)



```

## Joe Mauer

```{r}
library(dplyr)
library(stringr)
library(tidymodels)

set.seed(123)

# --- SETTINGS (keep consistent with your pipeline) ---
train_cutoff_year <- 2005  # use whatever you used earlier
joe_id <- "mauerjo01"

# --- 1) Pull Joe's row (as a prediction-only case) ---
joe_row <- AQCS %>% filter(playerID == joe_id)

# Safety checks
stopifnot(nrow(joe_row) == 1)

# --- 2) Rebuild the labeled modeling pool WITHOUT Joe ---
AQCS_model_mauer <- AQCS %>%
  filter(!is.na(hof)) %>%           # labeled only
  filter(playerID != joe_id)        # remove Joe entirely

# --- 3) Build the feature list exactly like your model used ---
model_features <- c(
  "WAR","PA","G","wRC+","Def","Fielding","Positional",
  "all_star_selections","gg","ss","mvp_awards","last_mlb_year"
)
model_features <- model_features[model_features %in% names(AQCS_model_mauer)]

# Assemble modeling table
model_df <- AQCS_model_mauer %>%
  select(all_of(model_features), hof) %>%
  mutate(hof = factor(hof, levels = c(0, 1)))

# --- 4) Modern-first split (WITHOUT Joe) ---
train_df <- model_df %>% filter(last_mlb_year <= train_cutoff_year) %>% select(-last_mlb_year)
test_df  <- model_df %>% filter(last_mlb_year >  train_cutoff_year) %>% select(-last_mlb_year)

# --- 5) Recipe + model + workflow (ridge logistic) ---
rec <- recipe(hof ~ ., data = train_df) %>%
  step_impute_median(all_predictors()) %>%
  step_normalize(all_predictors())

mod <- logistic_reg(penalty = tune(), mixture = 0) %>%
  set_engine("glmnet")

wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(mod)

# --- 6) CV tuning (WITHOUT Joe) ---
folds <- vfold_cv(train_df, v = 5, strata = hof)

grid <- grid_regular(
  penalty(range = c(-4, 1)),  # 1e-4 to 10
  levels = 20
)

tuned <- tune_grid(
  wf,
  resamples = folds,
  grid = grid,
  metrics = metric_set(roc_auc)
)

best_pen <- select_best(tuned, metric = "roc_auc")

# --- 7) Fit final model (WITHOUT Joe) ---
final_wf  <- finalize_workflow(wf, best_pen)
final_fit <- fit(final_wf, data = train_df)

# --- 8) Predict Joe's probability (Joe never touched tuning/training) ---
joe_pred_data <- joe_row %>%
  select(any_of(setdiff(model_features, "last_mlb_year")))  # predictors only

joe_prob <- predict(final_fit, new_data = joe_pred_data, type = "prob") %>%
  rename(p_hof = .pred_1) %>%
  mutate(name = joe_row$name, playerID = joe_id) %>%
  select(name, playerID, p_hof)

joe_prob

```

```{r}
All_Qualified_Catchers_Stats %>%
  arrange(desc
          (Def))

```
